---
title: "Introduction to R Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  In this tutorial, we will learn about R and how to set up R and RStudio on 
  your computer. We will also demonstrate basic R functionality and how to 
  install R packages from CRAN. 
---


```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
```

### Outline

In this tutorial, we learn about R and help you set up your computer to use R.

-   Why R?
-   What can R do?
-   Install R on your computer
-   Install the RStudio IDE
-   Install the `tidyverse` R package

### Why R?

R is a *free* and open-source programming language for statistical computing and graphics. The R Project for Statistical Computing (https://www.r-project.org) supports and develops R. 

**Why learn R?**
- R is free and open-source. Becasue it's free, everyone has access to it! Because it is open-source, anyone can add functionality to R in the form of packages. Also, statisticians and computer scientist can can audit the code and edit any bugs (i.e., there is no black box). 
- R is the lingua franca of statistics. In other words, it is the most common data analysis tool. Since R is the main computing environment for statistical analysis, both established and new methods in statistics, data mining, machine learning, and psychometrics are often employed using R, so there is no need for specialized software. Just learn R and you are ready to go. 
- R has beautiful data visualization capabilities. We will learn the basics of making graphs in R in this microcredential. 
- R has a great community! There are many talented R programmers that have contributed helpful R packages. In addition, there are great R mailing lists and Stack Overflow (https://stackoverflow.com) where you can find help using R. 


**R you ready?** 
The R community is made up of people passionate about the intersection of numbers, data, analysis, and code. In this first section, you will be introduced to some basic R syntax and discuss how R classifies data types so that it can mathematically process them in analysis.


*Background Information:* 
Unlike with other languages, most beginners who want to learn R do so because they want to analyze data. In this way, R is more of a tool to understand data than a programming language used to build software applications. However, what we are learning will build the foundation for you to start programming in R. 

**What is the tool for?**
R is powerful for conducting statistics and other specialized data analysis. It was invented by scientists for statistical computing and a community of specialized packages has been built around the language.


### What Can R do?
Let’s start with the basic syntax for mathematical calculations in R. R performs addition, subtraction, multiplication, and division with +, -, *, and / operators. 

Run the following code.

```{r  math1, exercise = TRUE}
573 - 74 + 1

25*2

10/5
```

Mathematical operations in R follow the standard mathematical order of operations. Let’s write your first line of R code and calculate some basic math! Solve 3+10/2.
```{r  math2, exercise = TRUE, exercise.eval = TRUE}

```
```{r  math2-solution}
3+10/2
```

Comments
Text written in a program but not run by the computer is called a comment. R interprets anything after a # as a comment.


Data Types 
Now that you know how to calculate basic math and add comments explaining your code, let’s dive into how R “thinks about” different types of data. In R and in programming, data types are the classifications we give different kinds of information pieces. In this lesson, we will explore the following R data types:

Numeric: Any number with or without a decimal point: 23, 0.03and the numeric null value NA.
Character: Any grouping of characters on your keyboard (letters, numbers, spaces, symbols, etc.) or text. Most strings are surrounded by single quotes: ' ... ' or double quotes " ... ", though we prefer single quotes. Sometimes you will hear this type referred to as “string.”
Logical: This data type only has two possible values— either TRUE or FALSE (without quotes). It’s helpful to think of logical types or booleans as on and off switches or as the answers to a “yes” or “no” question.
Vectors: A list of related data that is all the same type.
NA: This data type represents the absence of a value, and is represented by the keyword NA (without quotes) but it has its own significance in the context of the different types. That is there is a numeric NA, a character NA, and a logical NA.


Variables 
Now that you know how R classifies some of the basic information types, let’s figure out how to store them. In programming, variables allow us to store information and associate that information with a name. In R, we assign objectvariables by using the assignment operator, an arrow sign (<-) made with a less-than signcarat and a dash. 


Vectors 
We mentioned Vectors when we introduced data types earlier. In R, vectors are a list-like structure that contain items of the same data type.

Take a look here:

<Code Block>
spring_months <- c("March", "April","May","June")






## Install R

### How to install R

![](https://vimeo.com/203516510)

### Test your knowledge

```{r quiz1, echo=FALSE}
quiz(caption = "Quiz - Installing R",
  question("Is R free to download and use?",
    answer("Yes!", correct = TRUE, message = "R is free _and_ open source, which means that anyone can dowload _and_ use R."),
    answer("No.")
  )
)
```

## Install RStudio

### How to install RStudio

RStudio is an Integrated Development Environment for R. What does that mean? Well, if you think of R as a language, which it is, you can think of RStudio as a program that helps you write and work in the language. RStudio makes programming in R much easier and I suggest that you use it!

![](https://vimeo.com/203516968)

### Test your knowledge

```{r quiz3, echo=FALSE}
quiz(caption = "Quiz - Install RStudio",
  question("What is the RStudio?",
    answer("An application that makes it easier to use R.", correct = TRUE, message = "RStudio organizes your workspace and makes it easier to write, use, debug, and save R code. I highly recommend it and will rely on the RStudio IDE in several tutorials."),
    answer("An application that let's you use R without writing any code", message = "And thank goodness! Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel."),
    answer("Another name for R", message = "R and RStudio are two separate things. R is a language, like English. Think of RStudio as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    allow_retry = TRUE
  ),
  question("Is the RStudio free to download and use?",
    answer("Yes!", correct = TRUE, message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio with enhanced features, but we will not rely on those features in these tutorials."),
    answer("No.", message = "Like R, RStudio is free and open-source. There do exist professional versions of RStudio with enhanced features, but we will not rely on those features in these tutorials.")
  ),
  question("Do you need to install R if you already have RStudio?",
    answer("Yes.", correct = TRUE),
    answer("No.", message = "R does not come with RStudio; you need to install R separately.")
  )
)
```

## Install Packages

### How to install R packages

![](https://vimeo.com/203516241)

### Test your knowledge

```{r names, echo = FALSE}
quiz(caption = "Quiz - Working with Packages",
  question("What command do you use to install packages?",
    answer("`library()`", message = "We will see what `library()` does later."),
    answer("`install.packages()`", correct = TRUE),
    answer("`install_packages()`"),
    answer("There is no command. You must visit [cran.r-project.org](http://cran.r-project.org) and download packages manually.", message = "R makes it easy to download packages. You connect to the internet and then run one of the commands above."),
    allow_retry = TRUE
  ),
  question("How often do you need to install a package on your computer?",
    answer("Every time you restart R"),
    answer("Every time you restart your computer"),
    answer("Only once. Afterwards, R can find it on your hard drive as needed.", correct = TRUE),
    answer("Never, as long as you are connected to the internet.", message = "This could be true if you are using R over a cloud service. However, if you are using R locally on your own computer, you will need to install each package that you use on your computer."),
    allow_retry = TRUE
  ),
  question("What is the tidyverse?",
    answer("A collection of packages that work well together and provide tools for common data science tasks.", correct = TRUE, message = 'You can install all of the tidyverse packages at once with `install.packages("tidyverse")`.'),
    answer("A word processing program for R"),
    answer("A code style that reads like poetry"),
    answer("An alternate universe where everyone is an R programmer."),
    allow_retry = TRUE
  )
)
```


## Welcome

This is a demo tutorial. Compare it to the [source code](https://github.com/rstudio/learnr/tree/main/inst/tutorials/ex-setup-r/ex-setup-r.Rmd) that made it.

## Topic 1

### Exercise

*Here's a simple exercise with an empty code chunk provided for entering the answer.*

Write the R code required to add two plus two:

```{r two-plus-two, exercise=TRUE}
2+2
```

### Exercise with Code

*Here's an exercise with some prepopulated code as well as `exercise.lines = 5` to provide a bit more initial room to work.*

Now write a function that adds any two numbers and then call it:

```{r add-function, exercise=TRUE, exercise.lines = 5}
add <- function() {
  
}
```

## Topic 2

### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```
### Is this tutorial for you?

Do you need to work through the tutorial? Take the quiz below to find out.


question("Check all that you have NOT done:",
  answer("installed R on my computer", message = "* Install R"),
  answer("installed the RStudio IDE", message = "* Install RStudio IDE"),
  answer("installed the tidyverse R package", message = "* Install Packages"),
  answer("None of the above. I've done them all.", correct = TRUE, message = "You can skip this tutorial!"),
  type = "multiple",
  incorrect = "This tutorial is here to help! To get set up read:"
  ,
  question("Where do you download R?",
    answer("www.rstudio.com/download"),
    answer("[cloud.r-project.org](http://cloud.r-project.org)", correct = TRUE, message = "You can also download R from [cran.r-project.org](http://cran.r-project.org)"),
    answer("www.r-project.org", message = "Good try, but not exactly. www.r-project.org doesn't provide a download link, but it does provide a link to one the websites above."),
    answer("www.r.com"),
    allow_retry = TRUE
  ),
  question("How often should you update R?",
    answer("Everytime you use it", message = "This will be too often unless you use R very rarely!"),
    answer("About once a year", correct = TRUE, "A new version of R is released about once a year. Update sooner if you encounter a bug that you cannot explain."),
    answer("Never", message = "A new version of R is released about once a year. I'll assume that you are using the newest version of R, which will be the fastest version with the fewest unexpected behaviors." ),
    allow_retry = TRUE
  )
)



