---
title: "Introduction to R Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  In this tutorial, we will learn about R and how to set up R and the Positron IDE on 
  your computer. We will also demonstrate basic R functionality and how to 
  install R packages from CRAN. 
---


```{css, echo=FALSE}
.notice {
  padding: 15px;
  background-color: #f9f9f9;
  border-left: 6px solid #ffcc00;
  margin-bottom: 15px;
}


```{r setup, include=FALSE}
library(learnr)
tutorial_options(exercise.timelimit = 60)
gradethis::gradethis_setup()
```

## Outline

In this tutorial, we learn about R and help you set up your computer to use R.

-   Why R?
-   What can R do?
-   Install R on your computer
-   Install the Positron IDE

## Why R?

R is a *free* and open-source programming language for statistical computing and graphics.
The R Project for Statistical Computing (<https://www.r-project.org>) supports and develops R.

**Why learn R?**

-   R is free and open-source. Because it is free, everyone has access to it! Because it is open-source, anyone can add functionality to R in the form of packages. Also, statisticians and computer scientist can audit the code and edit any bugs (i.e., there is no black box).
-   R is the *lingua franca* of statistics. In other words, it is the most common data analysis tool. Since R is the main computing environment for statistical analysis, both established and new methods in statistics, data mining, machine learning, and psychometrics are often employed using R, so there is no need for specialized software. Just learn R and you are ready to go.
-   R has a great community! There are many talented R programmers that have contributed helpful R packages. In addition, there are great R mailing lists and Stack Overflow (<https://stackoverflow.com>) where you can find help using R.

**R you ready?**

The R community is made up of people passionate about the intersection of numbers, data, analysis, and code.
It was invented by scientists for statistical computing and a community of specialized packages has been built around the language.
In this first section, you will be introduced to some basic R syntax and discuss how R classifies data types so that it can mathematically process them in analysis.

*Background Information:* Unlike with other programming languages, most beginners who want to learn R do so because they want to analyze data.
In this way, R is more of a tool to understand data than a programming language used to build software applications.
However, what we are learning will build the foundation for you to start programming in R.

### What Can R do?

**Mathematical calculations.** Let's start with the basic syntax for mathematical calculations in R.
R performs addition, subtraction, multiplication, and division with +, -, \*, and / operators.

Run the following code using the blue "Run Code" button on the right.

```{r  math1, exercise = TRUE}
573 - 74 + 1

25*2

10/5
```

Mathematical operations in R follow the standard mathematical order of operations.
Let's write your first line of R code and calculate some basic math!
Solve 3+10/2 by writing code below and pressing "Run Code". The answer should be 8 (i.e., 10/2 = 5 and 3+5 = 8). Remember with order of operations division comes before addition. If you are not familiar with the order of operation or need a refresher, click [here](https://en.wikipedia.org/wiki/Order_of_operations).

```{r  math2, exercise = TRUE, exercise.eval = TRUE}

```

```{r  math2-solution}
3+10/2
```

<br> **Comments.** It's often good practice to annotate the code you write and not let the computer read it as code.
We can annotate or write text in our coding scripts using comments.
Comments are text written after a \# in a program that is not run by the computer.
R interprets anything after a \# as a comment.
A brief example is provided below demonstrating the use of comments.

```{r  comment, exercise = TRUE}
12 + 21 #this line of code sums 12 and 21

```

As you can see, the sum of 12 and 21 is shown when you run the code, but the comment is not displayed in the output.

**Data Types.** Now that you know how to calculate basic math and add comments explaining your code, let's dive into how R "thinks about" different types of data.
In R and in programming in general, data types are the classifications we give different pieces of information.
In this lesson, we will explore the following R data types:

-   **Numeric**: Any number with or without a decimal point: 23, -0.03 and the numeric null value `NA`.
-   **Character**: Any grouping of characters on your keyboard (letters, numbers, spaces, symbols, etc.) or text. Most strings are surrounded by single quotes: ' ... ' or double quotes " ... ", though we prefer single quotes. Sometimes you will hear this type referred to as a "string".
-   **Logical**: This data type only has two possible values--- either `TRUE` or `FALSE` (without quotes). It's helpful to think of logical types, or Boolean values, as on and off switches or as the answers to a "yes" or "no" question.
-   **Factor**: Used to represent groups or categories. R internally stores factors as an integer with an associated label. We will discuss factors in more detail later. 
-  `NA`: This data type represents the absence of a value, and is represented by the keyword `NA` (without quotes) but it has its own significance in the context of the different types. That is there is a numeric NA, a character NA, and a logical NA. There is also `NaN` which stands for "not a number"

**Variables and Objects.** Now that you know how R classifies some of the basic information types, we will almost always want to 'store' this information as a variable or as an object.
In programming, variables allow us to store information and associate that information with a name.
In R, we assign objects and variables by using the assignment operator, an arrow sign (\<-) made with a less-than sign and a dash.

Below is an example demonstrating how to assign data types to a variable or object.
In this case, we are assigning a student ID number to an object called `student_id`.
Run the first line of code written below.
You can see that R gives no output.
That's because you simply stored/assigned 98749 to the word `student_id`.
Now go ahead and type in `student_id` under the line of code written.
Take some time to think about the output R will give, if any at all.
Go ahead and run these two lines of code to show the output.

```{r  assignmentoperator, exercise = TRUE}
student_id <- 98749  #this assigns or 'stores' 98749 (a numeric data type) to the object/variable named 'student_id' 


```

<br> **Vectors.** In R, vectors are a list-like structure that contain items of the same data type. To combine the list of items to a vector, use the `c()` function and separate the items by a comma. 
In the example below, we assign a list of character data types (student ID numbers in this case) to a vector called `student_ids`. Notice the use of the assignment operator `<-` and the `c()` function to create the vector. *Note:* We could have made this a numeric vector by removing the quotes `"` from around the numbers (e.g., `student_ids <- c(98749, 042023, 052018, 061992)` ). 

Like the example above, if you run the line of code written below, R gives no output due to the same reason as above.
Take some time to think about the output R will give, if any at all, before typing in `student_ids` under the line of code written and pressing "Run Code."

```{r  vectorassign, exercise = TRUE}
student_ids <- c("98749", "042023","052018","061992") 


```

Try this yourself, create a character vector named `first_quarter` and store the months "January", "February", and "March" to the `first_quarter` vector. 

```{r  vectorassign2, exercise = TRUE, exercise.eval = TRUE}
 

```

```{r  vectorassign2-check}
gradethis::grade_result(gradethis::pass_if(~identical(.result, first_quarter <- c("January", "February", "March"))))
```

```{r  vectorassign2-hint-1}
first_quarter <- c() #finish the syntax
```

```{r  vectorassign2-hint-2}
first_quarter <- c("January",  ) # Don't forget to capitalize the names of the months!
```

```{r  vectorassign2-solution}
first_quarter <- c("January", "February", "March")
```

### Reflect

Let's take some time to reflect before moving on.
What is one thing that surprised you?
What is one thing that confused you?
Did you learn anything that might be useful in the type of work you do?

## Install R

Great!
Now that we know a little bit about R and what it can do, let's learn how to install the software onto your computer.

To be successful in this course, you will need to download and install two required software programs on your computer: R and Positron.
Don't worry, both are free!
The first software you will need to install is R which the the **programming language** we will be using in this course.
R is available for multiple computer operating systems such as Mac OS, Windows, and Linux.
Positron is the **software program** that makes R (the programming language) easy to use.
As a recap, R is the programming language we will be using and Positron is the software program/environment we can use R in.

We will first cover how to install R, the programming language, then we will cover how to install Positron.

You can navigate to the official R and Positron sites to download and install:

-[R](https://cran.rstudio.com/) 
<br>
-[Positron](https://positron.posit.co/download.html)

Please watch the video below on how to download and install R.

### Video: How to install R


```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 83.59133126934985%; height: 0;"><iframe src="https://www.loom.com/embed/8bd24ca9af17478aac9ca1ac9abc4695?sid=364c6a65-344c-456d-90c9-8f766ec539ab" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```


<br>
How to update your R installation without losing your packages: [Blogpost](https://www.r-bloggers.com/2015/06/how-to-upgrade-r-without-losing-your-packages/)

For more information and help installing R or Positron, please contact [UT Research Computing Support](https://oit.utk.edu/research/research-software/r-language/).


### Test your knowledge

```{r quiz1, echo=FALSE}
quiz(caption = "Quiz - Installing R",
  question("Is R free to download and use?",
    answer("Yes!", correct = TRUE, message = "R is free _and_ open source, which means that anyone can dowload _and_ use R."),
    answer("No.")
  )
)
```

## Install Positron

Great!
Now that you have successfully downloaded and installed R, let's walk through how to download and install Positron onto your computer.

### How to Install Positron

As we briefly described, Positron is an Integrated Development Environment for R.
What does that mean?
Well, if you think of R as a language, which it is, you can think of Positron as a program that helps you write and work in the language.
Positron makes programming in R much easier, and we suggest that you use it!

You can navigate to the official Positron site to download and install [Positron](https://positron.posit.co/download.html)

Please watch the video below on how to download and install Positron.

```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 83.20493066255777%; height: 0;"><iframe src="https://www.loom.com/embed/04f9d7275ca74224ad958787e759d4fb?sid=bc822164-a97f-4e91-bb95-23c21ee7c296" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```

For more information and help installing R or Positron, please contact [UT Research Computing Support](https://oit.utk.edu/research/research-software/r-language/).

### Test your knowledge

```{r quiz3, echo=FALSE}
quiz(caption = "Quiz - Install Positron",
  question("What is Positron?",
    answer("An application that makes it easier to use R.", correct = TRUE, message = "Positron organizes your workspace and makes it easier to write, use, and save R code. We will rely on the Positron IDE in several tutorials."),
    answer("An application that let's you use R without writing any code", message = "No, you'll still need to write code. And thank goodness! Code provides a reproducible record of your work, which is essential for data science."),
    answer("A spreadsheet program like Microsoft Excel.", message = "No, Positron is not a spreadsheet program like Microsoft Excel"),
    answer("Another name for R", message = "R and Positron are two separate things. R is a language, like English. Think of Positron as a program that helps you use the language, kind of like how a word processing program helps you write in English."),
    allow_retry = TRUE
  ),
  question("Is Positron free to download and use?",
    answer("Yes!", correct = TRUE, message = "Like R, Positron is free and open-source. There do exist professional versions of Positron with enhanced features, but we will not rely on those features in these tutorials."),
    answer("No.", message = "Like R, Positron is free and open-source. There do exist professional versions of Positron with enhanced features, but we will not rely on those features in these tutorials.")
  ),
  question("Do you need to install R if you already have Positron?",
    answer("Yes.", correct = TRUE),
    answer("No.", message = "R does not come with Positron; you need to install R separately.")
  )
)
```

## Getting Started with R and Positron

Great!
Now that you have successfully installed R and Positron onto your computer.
Let's take a little tour to get you familiar with the software.

## Positron Tour and Customization

The videos below will help you familiarize yourself with and customize Positron.

### R Sessions and R Scripts

First, let's learn how to create an R session and R script in Positron.
<br>
```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 77.25321888412017%; height: 0;"><iframe src="https://www.loom.com/embed/1868ee3cc041488ebbbb8d3399500ba1?sid=4894f43c-9ba0-4850-b44c-7844b024e274" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```
<br>

### Themes and Extensions
Next, let's customize Positron by choosing a theme and adding extensions.
<br>
```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 76.59574468085107%; height: 0;"><iframe src="https://www.loom.com/embed/5e545d0201db47b6b36b27e6c9efff22?sid=95fb8167-e7f3-477f-bb2a-c3471d492924" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```
<br>

### Positron Tour
Finally, let's take a quick tour of the Positron IDE and its many features.
<br>
```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 77.80979827089337%; height: 0;"><iframe src="https://www.loom.com/embed/30fad0539acf4e5f872647ab5b858a51?sid=5110ac65-8671-4a85-8730-7dc8fc2dcdf6" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```
<br>

## Working Directories

We hope you enjoyed our virtual tour of Positron.
The next thing we want to cover are working directories and folders.
Both let Positron know where on your computer you're saving your files or opening files from.
These are important to use properly when creating R files/scripts (otherwise you will get frustrating errors messages).

The video below will give you a brief tutorial of setting your file location using a working directory or folder.
<br>
```{r, results='asis', echo=FALSE}
cat('
<div style="position: relative; padding-bottom: 78.14761215629522%; height: 0;"><iframe src="https://www.loom.com/embed/27b922cca43d4407bd7ae9b906ca2776?sid=0121362b-9e44-49f9-a097-243057788f93" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 100%; height: 100%;"></iframe></div>
')
```
<br>

## Tasks

Now that you have R and Positron installed on your machine and you have been acquainted with Positron, we have provided a few tasks below for you to complete.

1.  Choose a preferred appearance theme for Positron
2.  Create one new R script which assigns the following numbers to an object called `student_ids`: 071194, 101022, 042523, 052523.
3.  Save the R script in the weekly folder for this course. Please name the script the following: lastname-R-task-tutorial-1. Upload it to Canvas.

### Reflect

Let's take some time to reflect before moving on.
What is one thing that surprised you?
What is one thing that confused you?
Did you learn anything that might be useful in the type of work you do?

### Wrap-up

Good job on completing the first of three tutorials within the first module of this course!
Please move onto the next tutorial where we will learn foundational R and data analytic skills.