---
title: "Nuts and Bolts of Data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.eval = TRUE)
library(tidyverse)
gradethis::gradethis_setup()
```


## Introduction

Welcome! Now that you understand how to use packages, functions, and how to import data into R, we can start to get familiar with basics of data in a data science sense. The purpose of this tutorial is to get you familiar with data structures, data types, and accessing data in R. 

This tutorial will cover the nuts and bolts of data within R in terms of knowing the different data types and structures and understanding how to access specific elements within a data object. These concepts are fundamental and essential in gearing us up to learn about data manipulation in the next module. We will be covering the following in this tutorial:

- Data types
- Data structure
- Data indexing

## Data Types

Data types are the types of data supported and can be used with R and within RStudio. While there are different data types that can be used within R, we will cover the 4 basic data types:

- Numeric
- Integer
- Character
- Logical

### Numeric

Numeric data types are real number values that are allowed to have decimal points and are not considered whole numbers.
```{r  numerictypes, echo=TRUE}
# Numeric atomic vector
grade1 <- 93.51
typeof(grade1)
str(grade1)

```

In this example above, we are saving the numerical value 93.5 to an object called `grade1`. When you run the line of code `grade1`, you can see it will give you the value of 93.5. Additionally, we can use the `typeof()` function to check the data type in which the output says "double" which is true since all numeric data types are considered double numerical values since they are allowed to have decimal points. We can also use `str()` function to check the structure of the object `grade1` which the output says it's `num`, a numerical data type, with the value 93.5 inside of the object.

While we used 93.5 to demonstrate a numeric data type, 93 is also considered a numeric data type which is synonymous as a double numerical value even though this value does not have specified decimal places. Check out the example below :smiley:

```{r  numeric_types3, echo=TRUE}
# Numeric atomic vector
grade2 <- 93
grade2
typeof(grade2)
str(grade2)
```

You're probably wondering why 93 is considered a double numerical value and not an whole number integer. That's a good curiosity to have which leads us to discuss integer values.

### Integer

Integer data types are real numbers that are not allowed to have decimal points as opposed to numeric data types. Specifically, the suffix `L` is used to specify integer data.

```{r  integer_types, echo=TRUE}
integer <- 256L
integer
typeof(integer)
str(integer)
```

As you can see when we use `typeof()` and `str()`, the output tells us that the object `integer` is an integer data type. You will seldom use integer data types in your work (probably never), however, we wanted to provide you a brief review of this data type.

### Character

Character data types are aptly named as they are data types that are used to specify data that are characters and string values. String values are a set of character data while a character value is a single character. For example, `'T'` is considered a single character while `"Theo"` is considered a string. Going one more step in this distinction `"Hi my name is Theo"` is also considered a string of character values. Strings can be represented through the use of single quotes `''` or double quotes `""`. As a common rule, `''` is used for character variables and `""` is used for string variables.

```{r  char_types, echo=TRUE}

name <- "Theo"
name
typeof(name)
str(name)
```

When we use `typeof()` and `str()`, the output tells us that the object `name` is an character data type (string in this case).

### Logical

Yay!We made it to the final data type: logical data. Logical data types in R are known as boolean data types. Logical data can only have two values: `TRUE` and `FALSE`.

```{r  logical_char, echo=TRUE}

logical1 <- TRUE
logical1
typeof(logical1)
str(logical1)

logical2 <- FALSE
logical2
typeof(logical2)
str(logical2)
```

Additionally, logical data types can also be represented as `T` and `F` which is shorthand for `TRUE` and `FALSE`. R accepts both form of logical data types, however, it's standard practice to use `TRUE` and `FALSE` rather than `T` and `F`.

## Data Structures

Now that you know about the common data types, we can talk about how we can store these different types of data in R which leads us to talk about data structures.

A data structure is a format in how data is organized and stored in order to be efficiently accessed. Our focus will be on the base data structures within R: vectors, arrays, lists, and data frames. There are two attributes that define each of these data structures: 1) dimensionality and 2) content type. 

Data structures can be organized by dimensionality (they are either 1-dimensional, 2-dimensional, N-dimensional (e.g., however many dimensions you want to make the structure) and whether or not the content stored inside the structures are homogeneous (i.e., all data contents must be the same type) or heterogeneous (i.e., all contents can be different types)). Based on these two attributes that define data structures, there are five common data structures used within R and data science in general. Table 1 outlines these five different types and classifies their two attributes.

```{r echo=FALSE, results='asis'}
library(knitr)
data_structure <- data.frame(
  Dimension = c("1d", "2d", "nd"),
  Homogenous = c("Atomic vector", "Matrix", "Array"),
  Heterogenous = c("List", "Data frame", " ")
)
kable(data_structure, caption = "Table 1: Data Structures")
```

Here are the 5 common data structures outlined below:

- Atomic vector: Homogeneous and 1-dimensional
- List: Heterogeneous and 1-dimensional
- Matrix: Homogeneous and 2-dimensional
- Array: Homogeneous and =< 2-dimensional
- Data frame: Heterogeneous and 2-dimensional

### Atomic vector

There are four common types of atomic vectors: numeric, integer, logical, and character. 

#### Numeric atomic vectors

Numeric atomic vectors can handle numerical values that are not considered whole numbers.Go ahead and use the exercise below to make a numeric vector.
```{r  num_vectors, exercise = TRUE}
# Numeric atomic vector
num_vector <- c(1.5, 9.7, 2)
num_vector
```
Note: For the lines of code in the exercise above, remember that the arrow "<-" is assigning 'something' to an object. In this case, we are assigning/storing a vector to an object called `num_vector`. When we only run the line of code with the object `num_vector` we are able to see/access the stored data inside of the object which is a vector in this case.

Another note: Whenever we want to make an object that contains more than one value, we have to use the ``c()` and separate each value with a comma. The `c()` stands for combine. So in the example above, we are combining the numerical values of 1.4, 9.7, and 2 and assigning/storing it to an object (which in this case is a vector) called `num_vector`.

#### Integer atomic vectors

Integer atomic vectors can handle numerical values that are considered whole integers. The `L` after each numerical value provided in the exercise below tells R that these values are integers and not numeric. Go ahead and use the exercise below to create an atomic vector that is storing integers. What do you think your output will be when you run the `int_vector` line of code?
```{r  int_vectors, exercise = TRUE}
# Integer atomic vector
int_vector <- c(2L, 5L, 9L)
int_vector
```

#### Logical atomic vectors

Logical atomic vectors can handle logical values. In R (and other programming languages), logical vectors are expressed by `TRUE`, `FALSE`, `T`, and `F`. Logical values will come in handy when we talk about logical comparisons later in this tutorial :-) Go ahead and use the exercise below to create a logical vector.
```{r  log_vector, exercise = TRUE}
# Logical atomic vector
log_vector <- c(TRUE, FALSE, T, F)
log_vector
```

#### Character atomic vectors

Lastly, character atomic vectors can handle character values in R. Character values can be anything that is wrapped in quotations (this is also known as a string). Go ahead and use the exercise below to create a character vectors with four unique strings stores inside (in this case, these are names of schools). 
```{r  char_vector, exercise = TRUE}
# Character atomic vector
char_vector <- c("Bell Middle", "Kim High", "South Elementary")
char_vector
```

#### Atomic vector overview

As you can see in the different examples provided above, we created four different types of atomic vectors that were 1-dimensional (a single row) and each stores the same data type (homogeneous).

### Lists

Lists are like atomic vectors but allow elements inside of the list to be heterogeneous elements instead of homogeneous. We can make a list by using `list()` instead of `c()`. The exercise below provides an example of a list which contains different elements. We can check the structure of the list object by using `str()`.

```{r  list, exercise = TRUE}
list_example <- list(100:105, "apple", c(TRUE, FALSE, FALSE), c(99.8, 87.0, 88.3))
str(list_example)

```

Let's walk through the code in the exercise. First, we are using the function `list()` to create an object that's a list. Within this function, we are providing different arguments to tell R what to include in the list. The first argument given `100:105` tells R to create the first element in the list which includes numbers from 100 to 105. The `:` notation between 100 and 105 tells R to include values between those two numbers, thus providing the following output for the first element within this list: 100, 101, 102, 103, 104, 105. The second argument tells R to create the second element in this list which includes the character type `apple`. The third argument tells R to create the third element in this list which are logical values. The last argument tells R to create the fourth element in this list which includes 99.8, 87.0, and 88.3.

When we use the `str()` function to check the structure of the list, the output tells us that the object `list_example` is a list that contains 4 elements and provides us with the data types for each of these elements as well as the contents inside of each element. 

### Matrix

Matrix are similar to atomic vectors is that they only allow elements to be homogeneous but they are 2-dimensional objects rather than 1-dimensional. We can make a matrix by using `matrix()` and providing the elements to be included in the matrix and by specifying the dimensions of the matrix by using `ncol` and `nrow`. The exercise below provides an example of a matrix which contains all numeric elements. 

```{r  matrix, exercise = TRUE}
matrix_example <- matrix(1:9, ncol=3, nrow = 3)
matrix_example

```

In walking through the code, we are using the function `matrix()` to create an object that's a matrix. The first argument provided in the function, we are telling R that we want to include numbers 1 through 9 in the matrix (all numeric) and that we want to have 3 columns and 3 rows. So when we run `matrix_example` on the second line, you can see that a matrix is produced with the values 1 through 9 and is organized by 3 columns and 3 rows. 

### Array

Arrays are similar to a matrix (or matrices) in the sense that they store homogeneous elements but they allowed to have more than 2 dimensions if needed/necessary. We can make an array by using `array()`. The exercise below provides an example of an array with 3 dimensions.

```{r  array, exercise = TRUE}
#  4 rows,2 columns and 3 arrays
array_example <- array(c(1,2,3,4,5,6, 60, 18, 12, 13, 14, 19), dim = c(4,2,3))
array_example

```

When walking through the code, we are using the function `array()` to create an object that's an array. The first argument provided in the function, we are giving R the elements we want displayed in the array. These are all numeric (homogeneous) elements. The second argument provided in the function `array()` is the dimensions needed for the array to be created (e.g., `dim =`). So for `dim = c(4, 2, 3)` we are saying we want each dimension of the array to have 4 rows, 2 columns, and 3 dimensions. When you run the line of code `array_example` you can see that three dimensions are provided (matrices) and each of these have 4 rows and 2 columns. The placement of the numerical elements `c(1,2,3,4,5,6, 60, 18, 12, 13, 14, 19)` starts in the first matrix (`, , 1`) at the first row and first column and continues to place the numerical elements provided in a row wise manner until all numerical elements have been listed (the last value (19) provided in the second matrix in the fourth row, first column. But in order to continue to build the array as specified by the dimensions given in the `dim` argument, R continues to starts to loop the numerical elements provided by starting again at placing 1 in the second matrix in the first row, second column and continues on until the full array is created as specified.

### Data frames

Yay!:tada: We made it to the final data structure: data frames. Data frames will more than likely be the most common data structure you will use in the education research field, and is the most common way to store data in R. Data frames are dimensional (rows and columns) and can store heterogeneous elements. For data frames, columns are typically variables and rows are cases/observations. We can make a data frame from scratch using `data.frame()`, however, more often than not, you will have a data file (e.g., .csv) and will read it into R (much like you did in the past mini module when reading in the hsb data set.). Regardless, let's make a data frame in the example below. 

```{r  dataframe, exercise = TRUE}
# pass these vectors as input to the array.
#  4 rows,2 columns and 3 arrays
class_data <- data.frame (studentid  = c(101, 102, 104, 105, 106),
                  gender = c("Female", "Female", "Male", "Female", "Male"),
                  test1 = c(98, 82, 85, 97, 92),
                  mathmajor = c(1, 0, 0, 1, 1)
                  )
class_data

```

In creating the data frame, we specify that the first column will be student ids, second: gender, third: test 1 score, and fourth: if the student is a math major. As you can see when you run the line of code that read `class_data` the data frame has two dimensions (rows and columns) and has heterogeneous elements (numerical and character values). 


## Data Indexing

Now that we know about data types and data structures, we can now talk about how to access specific values/elements inside of a data structure (this is also referred to as data indexing).

### Indexing for Vectors

Vectors are 1 dimensional so if we need to access a certain value/element inside a vector, we need to specify the location of the value we want to extract. Let's look at the same example we used for character vectors earlier. 

```{r  index_vec, exercise = TRUE}
char_vector <- c("Bell Middle", "Kim High", "South Elementary")
char_vector

```

Let's say we want to save the second school name into another object called `hs`. We can achieve this by specifying the location of the second element within `char_vector` using `[[]]`.

```{r  index_vec2, exercise=TRUE, exercise.setup = "index_vec"}
hs <- char_vector[[2]]
hs
```

As you can see in the example above, we access the second element inside of `char_vector` through this code `char_vector[[2]]` to which we save it to another object called `hs`.

After saving Kim High to another object called `hs`, do you think Kim High still exists as the second element in `char_vector`? Let's check!

```{r  index_vec3, echo=TRUE}
char_vector <- c("Bell Middle", "Kim High", "South Elementary")
char_vector
hs <- char_vector[[2]]
hs
char_vector
```

### Indexing for Lists

Let's go ahead and try to access certain values in a list! We're going to use the same example above when we learned about lists. Let's access the different first two data types specified in this list below.

```{r  index_list, echo=TRUE}
list_example <- list(100:105, "apple", c(TRUE, FALSE, FALSE), c(99.8, 87.0, 88.3))
list_example
list_example[[1]]
list_example[[2]]
```

Great! We were able to access the first element of the list that contained the values 100 - 105 and the second element of the list that contained the string `"apple"`. Let's try accessing one value in this list. Specifically, let's save the value 103 to a new object called `quiz1`.

```{r  index_list2, echo=TRUE}
list_example <- list(100:105, "apple", c(TRUE, FALSE, FALSE), c(99.8, 87.0, 88.3))
list_example
quiz1 <- list_example[[1]][4]
quiz1
```

In the code `list_example[[1]][4]` we tell R to access the first element of `list_example` which is a set of numerical values by using `[[1]]` then we tell R to grab the fourth value within the first element (set of numerical values 100 - 105) by using `[4]`.

### Indexing for Matrices

Great! So far we have only accessed values in 1-dimensional data structures so let's try accessing values in two dimensional data structures. Specifically, let's use the example below to access all values within column 2

```{r  index_matrix, echo=TRUE}
matrix_example <- matrix(1:9, ncol=3, nrow = 3)
matrix_example
matrix_example[, 2]

```

Good job! Let's use the same example but access all values in the second row.

```{r  index_matrix2, exercise=TRUE}
matrix_example <- matrix(1:9, ncol=3, nrow = 3)
matrix_example
matrix_example[2, ]

```

How do you think you would access the value located in the second column and second row? Give it a try below.

```{r  index_matrix3, echo=TRUE}
matrix_example <- matrix(1:9, ncol=3, nrow = 3)
matrix_example


```


Here's the answer!

```{r  index_matrix4, echo=TRUE}
matrix_example <- matrix(1:9, ncol=3, nrow = 3)
matrix_example
matrix_example[2, 2]

```

### Indexing for Arrays

Since arrays are similar to matrices (the only difference is arrays can support more than 2 dimensions), we won't cover the specifics on indexing for arrays. Which is perfect since we want to dedicate more time and training to indexing and accessing different values within data frames since this is the data structure you will more than likely use the most.

### Indexing for Data Frames

Yay!:tada: We made it to the portion of this module where we will learn how to index and access data within data frames. Let's use the data frame we created earlier in the example below.

```{r echo=TRUE}
class_data <- data.frame (studentid  = c(101, 102, 104, 105, 106),
                  gender = c("Female", "Female", "Male", "Female", "Male"),
                  test1 = c(98, 82, 85, 97, 92),
                  mathmajor = c(1, 0, 0, 1, 1)
                  )
class_data
```

#### Accessing names of columns and rows

If we have a data frame, we can easily view the column names without opening up the data frame itself by using `names()` or `colnames()`. 

```{r echo=TRUE}
names(class_data)
colnames(class_data)

```

If needed, we can also get the rownames by using `rownames()`.

```{r echo=TRUE}
rownames(class_data)
```

Additionally, we can do a bit of manipulation to the data frame. The column names of the `class_data` data frame are already named, but what if we wanted to change the name of the colnames to `id`, `sex`, `test`, `major`? We can accomplish this by using `colnames()` and saving the dataframe back into the original data frame titled `class_data`.

```{r echo=TRUE}
colnames(class_data) <- c("id", "sex", "test", "major")
class_data
```

Please be careful if you alter data and save it back into the original data frame (like we did above) because the assignment operator `<-` can at times serve as a save function. If you wanted to change the colnames but save this updated data frame as a differently named object, we can achieve this by saving the changed data frame to different named data frame.

Let's take the recently altered `class_data` data frame and change the column names back to the original column names but save it as a data frame called `class_data_raw`.

```{r echo=TRUE}
class_data_raw <- class_data
colnames(class_data_raw) <- c("studentid", "gender", "test1", "mathmajor")
class_data_raw
class_data
```

In the code above, we saved the data frame `class_data` to another data frame called `class_data_raw`. At this point, the contents inside `class_data` are identical to `class_data_raw`. In the second line, we change the column names of `class_data_raw` to `studentid`, `gender`, `test1`, `mathmajor`. In the last two lines of code, we can see that `class_data_raw` has the original column names(`studentid`, `gender`, `test1`, `mathmajor`) and `class_data` has the updated column names (`id`, `sex`, `test`, `major`).

#### Accessing specific elements of a data frame

Since data frames are 2 dimensional structures, we can access specific element(s) within a data frame by specifying the rows and/or column.

Let's access the `test1` grade for `studentid` 104 using the data frame from our example.

```{r echo=TRUE}
class_data_raw[3, 3]
```

We can also access multiple values as well. Let's retrieve the `test1` grade for `studentid` 104 and 106.

```{r echo=TRUE}
class_data_raw[c(3, 5) , 3]
```

We can also access only the values within the `test1` column from the data frame through the use of indexing by only specifying the column location but not specifying a specific row.

```{r echo=TRUE}
class_data_raw[, 3]
```

Alternatively, if we know the name of the column that we want to access all values, we can access this column by using the `$` notation.

```{r echo=TRUE}
class_data_raw$test1
```

Note: R is case sensitive and needs the exact spelling in order to access specified rows. If you were to use `class_data_raw$Test1` you will either receive an error message or a `NULL` value as that column does not exist in the data frame.

## Special Data Type: Factors

Before we close out this sub module, there are two additional topics we wanted to cover that don't fit neatly into the basic framework outlined above when it comes to data considerations but will be largely helpful for when we get into data wrangling.

The first special topic is understanding factors.

When discussing data types earlier, we noted the following as the four basic data types:

*-* Numeric
*-* Integer
*-* Character
*-* Logical

When working with education-related data in R, you will also be working with a specific data type called factors. Factors are a special case of character variables that represent categorical variables. For example, you will be working with demographic data that often contains categorical variables such as sex or race/ethnicity. If you are working with a data set, you may have a sex categorical variable that may have a limited number of values (e.g., female, male). This is different from a data set with a variable (or column) of names as a column of names will have numerous different possibilities to where this variable would not be considered a categorical variables (or factor) and is instead just considered a character variable. 

We can create a factor variable using `factor()`.

```{r echo=TRUE}
classification <- factor(c("freshman", "sophomore", "junior", "senior"))
classification
```

We can check the structure of the data using `str()`

```{r echo=TRUE}
str(classification)
```

In the output above, you can see that R tells us that the object called `classification` is a factor with four levels. When you create factors in R, you automatically defines the factor levels based on alphabetical order. As you can see in the output, R is labeling `freshman` as level 1, `sophomore` as level 4, `junior` as level 2, and `senior` as level 3.

The ordering of levels in this case is not intuitive so let's go ahead and change it! We can reorder the levels of a factor using the `levels` argument in the `factor()` function.

```{r echo=TRUE}
classification <- factor(classification, levels = c("freshman", "sophomore", "junior", "senior"))
```

Let's check if the reordering was successful.

```{r echo=TRUE}
str(classification)
```

Yay, it worked! :)

### Convert character strings into factors

More often than not, when you import data into R, it won't automatically import categorical variables as factors but as strings instead. No worries! We can convert these column variables into factors.

To illustrate this, let's go ahead and make an object with strings called `school_type`.

```{r echo=TRUE}
school_type <- c("private", "public", "charter")
str(school_type)
```

We can convert these strings into factors using the `as.factor()` function and check by using the `str()` function.

```{r echo=TRUE}
school_type <- as.factor(school_type)
str(school_type)
```

Yay! Now you know about factors :)

## Special Use: Logical Comparisons

Now that we have covered factors, let's talk about logical operators and comparisons!

As noted above, we are now familiar with logical data types, but we haven't discussed logical operators and comparisons. This will come in handy once we get into more advanced data wrangling in the next module. 

The chart below shows a cheatsheet of possible logical comparisons that can occur in R.

```{r log-comp-chart}
knitr::include_graphics(paste0(getwd(), "/images/logicalchart.png"))

```

For example, if we want to know if two objects are equal we can use the following:

```{r echo=TRUE}
1 == 1
```

The output will return the logical operator `TRUE`.

Here's another example using values stored in an object:
```{r echo=TRUE}
object1 <- 106
object2 <- 509

object1 > object2
```

We can also use logical comparisons for character values!
```{r echo=TRUE}
"Bob" != "Bobby"
```

Lastly, we can compare values within two lists:
```{r echo=TRUE}
list1 <- c(906, "Juan", 1000, "Gracie")
list2 <- c(906, "juan", 0001, "Gracie")

list1 == list2
```
In the output we can see that the first element in `list1` and `list1` are equal values, the second element are not equal (remember, R is case sensitive), third element is not equal, and the fourth element is equal.

There's a lot we can do with logical comparisons but we wanted to introduce you to logical comparisons and operators as they will come in handy when we begin to wrangle our data in the next module.

## Tasks
Now that you are more acquainted with data types and structures, let's engage in a task! Please follow the steps outlined below.

1. Open the R script you created in the previous submodule (lastname-R-task2) that is saved in your project folder for this course. You should have code written to load the tidyverse and psych libraries and to read in your hsb dataset.Go ahead and run this script to make sure your code works (we are going to pick up where we left off last time).
2. Using  comments within R, write notes on the names for each of the variables in the hsb dataset.
3. Using comments, write notes on the types of data for each variable in the hsb dataset.
4. Ensure the following 4 variables are classified as factors: gender, race, ses, schtype, prog. Save these changes to another data frame object called `hsbclean`.
5. Change the following column names and save these changes to another data frame object called `hsbclean`.
- schtype -> schooltype
- prog -> program
- socst -> socialst
6. Save R script as lastname-R-task3
7. Read the brief resource in Canvas titled: datawrangling-r-cheat-sheet.pdf and ADD ANOTHER CHEATSHEET/RESOURCE

#### Note:
You may run into issues in completing the tasks listed above. Please refer to this document as well as the previous Learnr documents. Google is also a great resource when you run into issues in R.

### Reflect
Let's take some time to reflect before moving on. What is one thing that surprised you? What is one thing that confused you? Did you learn anything that might be useful in the type of work you do?

### El Fin
Good job on completing the first of three tutorials within the first module of this course! Please move onto the next module on Canvas where we will learn about data wrangling.
