---
title: "Statistical Concepts: An Overview"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r load, include=FALSE}
library(learnr)
library(tidyverse)
library(ggthemes)
tutorial_options(exercise.timelimit = 60)
tutorial_options(exercise.eval = TRUE)
```

## Introduction

Welcome back to the final module of this course. In the previous modules we learned how to import, transform, wrangle, and tidy data using the `tidyverse` package. In this final module, we will learn how to generate basic descriptive statistics.

```{r process, echo=FALSE, out.width="75%"}
knitr::include_graphics(paste0(getwd(), "/images/base.png"))

```

Before we delve into the functions for descriptive statistics in R, let's first introduce and review the following foundational statistics terminology: 

- Population vs. Sample
- Parameter vs. Statistic
- Dependent vs. Independent variables
- Levels of measurement
- Types of data
- Randomization
- Internal vs. External validity

Now that we know where we've been and where we're headed, let's jump into it!

## Population vs. Sample

The difference between a population and sample is one of the first concepts taught in statistics and research methods courses. A **population** is the complete group about which you want to make conclusions, and a **sample** is a subset of the population of interest. Researchers often collect data from a sample to make conclusions (or draw inferences) about the entire population of interest.

```{r tidychart, echo=FALSE, out.width="75%"}
knitr::include_graphics(paste0(getwd(), "/images/population.png"))

```

For example, all undergraduate students at the University of Tennessee (UT) would be considered the entire undergraduate population of UT. However, a random selection of 200 undergraduate students from UT would be considered a sample of the population.

```{r quiz1, echo = FALSE}
quiz(caption = "Population vs. Sample Quiz",
  question("What is a population in research?",
    answer("The complete group about which you want to make conclusions", correct = TRUE),
    answer("A subset of the population of interest"),
    answer("A group of individuals participating in a study"),
    answer("A group of individuals from different backgrounds")
  ),
  question("What is a sample in research?",
    answer("The complete group about which you want to make conclusions"),
    answer("A subset of the population of interest", correct = TRUE),
    answer("A group of individuals participating in a study"),
    answer("A group of individuals from different backgrounds")
  )
)

```

Now that we know the difference between a population and a sample, we can discuss the difference between a parameter and a statistic.

## Parameter vs. Statistic

**Parameters** are numbers that describe the properties of an entire population, and **statistics** are numbers that describe the properties of a sample. 

Let's turn back to our example of undergraduate students at UT. In [Fall 2023](https://irsa.utk.edu/reporting/fact-book/), 28,883 undergraduate students were enrolled at UT which represents the entire undergraduate population at UT. Let’s say we have the average age of all 28,883 undergraduate students, and the average age is 22 years old. The average age of the entire undergraduate student body at UT would be considered a *parameter* as it describes the average age of a population of interest. On the other hand, let’s say we didn’t have access to the age of all 28,883 undergraduate students but collected a sample of 200 students and asked how old they were. We averaged the age of the 200 student sample and found the average age of this sample to be 21.7 years old. The average age of this sample would be considered a *statistic* as it describes a properties of a sample of interest.

In most research cases, it is difficult to collect data for an entire population, so we instead take a sample in which we can make inferences about the population. 

```{r quiz2, echo = FALSE}
quiz(caption = "Parameter vs. Statistic Quiz",
  question("What is a parameter?",
    answer("A number that describes the properties of an entire population", correct = TRUE),
    answer("A number that describes the properties of a sample"),
    answer("A variable that is manipulated in an experiment"),
    answer("A variable that is measured in an experiment")
  ),
  question("What is a statistic?",
    answer("A number that describes the properties of an entire population"),
    answer("A number that describes the properties of a sample", correct = TRUE),
    answer("A variable that is manipulated in an experiment"),
    answer("A variable that is measured in an experiment")
  )
)

```


## Independent vs. Dependent Variables

Researchers often examine relationships between variables in a sample of interest. Variables are characteristics of interest in a sample or population that can take on different values. For example, variables can be height, age, test scores, etc. 

### Independent Variable

Researchers want to examine relationship between variables as guided by their specific research question. This examination is done by specifying an independent variable and a dependent variable. An **independent variable** is a variable that varies or can be manipulated to explore the effect or relationship it has with a **dependent variable**. As the name implies, independent variables are variables that are independent of other variables/characteristics. Meaning that no other variables in the research study impact the independent variable. 

Independent variables can be experimental, that is directly manipulated by the researcher (e.g., treatment conditions in an experiment), or naturally occurring characteristics of a subject such as height, weight, education, age. 

### Dependent Variable

A **dependent variable**, also called the *outcome variable*, is hypothesized to change as a result of changing or manipulating the **independent variable**. In other words, the dependent variable's changes are contingent upon the independent variable. The dependent variable is the outcome of interest. A researcher aims to examine how the dependent variable changes as the independent is manipulated or varied. 

### Example

A classic example of independent and dependent variables is examining the relationship between students' study hours and their test score. In other words, if students study more, do their test scores increase? If a researcher wishes to examine the relationship between study hours and test performance, the independent variable would be the amount of time a student studies and the dependent (or outcome) variable would be test scores. 

```{r quiz3, echo = FALSE}
quiz(caption = "Independent vs. Dependent Variables Quiz",
  question("What is an independent variable?",
    answer("A variable that varies or can be manipulated to explore the effect or relationship it has with a dependent variable", correct = TRUE),
    answer("A variable that is dependent on other variables"),
    answer("A variable that is manipulated in an experiment"),
    answer("A variable that is measured in an experiment")
  ),
  question("What is a dependent variable?",
    answer("A variable that varies or can be manipulated to explore the effect or relationship it has with an independent variable"),
    answer("A variable that is dependent on other variables"),
    answer("A variable that is manipulated in an experiment"),
    answer("A variable that is measured in an experiment", correct = TRUE)
  )
)

```

## Levels of Measurement

In statistics, variables are often quantified based four levels of measurement:

- Nominal
- Ordinal
- Interval 
- Ratio

The levels of measurement are listed in ascending order, starting with nominal variables, based on the progressively detailed information they provide. 

### Nominal Variables

Nominal variables are names of categories data can fall in. They are also known as *categorical data*. An easy way to remember what nominal variables are is to think *nominal* = *names*. For example, students’ major can be considered a nominal variable in a dataset with postsecondary student data (e.g., accounting major, statistics major, education major). 
Nominal variables do have a countable number of distinct groups so they cannot be unknown to the researcher or infinite. While nominal variables have a countable number of named categories, they don't have a natural order. Going back to our example of students' major, there is not natural order of accounting, statistics, or education majors as categories. Statistics is greater, equal, or less than the other majors. 

Other examples of nominal variables can be gender, race/ethnicity, college minors.

Binary variables (also called dichotomous variables) are a type of nominal variables which can only have two distinct values. Examples of binary variables are pass/fail, control/treatment.

We often set nominal variables as **factors** in R to ensure that the data is treated as categorical data.

### Ordinal Variables

Ordinal variables are similar to nominal variables since they too have categories of data, however, they provide more information than nominal variables, since ordinal variables can be ranked following a natural order of values.

Example of ordinal variables are education levels (primary, secondary, post-secondary), income (low, middle, high), class rank. 

While ordinal variables can be ranked, we don't know the exact degree of difference between these categories. Let's take class rank for example. Person A might have a class rank of 10, Person B a class rank of 11, and Person C a class rank of 12. While we know the rank of these individuals, we don't know the relative absolute different between these individuals. For example, Person A might have a GPA of 3.978, Person B a 3.969 GPA, and Person C a 3.900 GPA. While they're class rank is ordered, they don't have the same distance apart from each other in terms of GPA which is why we rank them if we don't know the relative degrees of difference between them. 

### Interval Variables

The next most descriptive level of measurement are variables measured on an interval scale. Interval scales are used for data that's continuous without a true zero. For example, SAT scores are measured on an interval scale where the distance between two points is meaningful (unlike ordinal variables) since we know the relative degrees of difference between these two points. However, SAT scores don't have a true zero point (or 0 does not represent the absense of the trait being measured).


### Ratio Variables

The most descriptive level of measurement are ratio variables. Like interval variables, ratio variables are also used for continuous data, but are different for having a true zero measurement which represents a lack of an attibute. For example, height, weight, speed, and time are all ratio variables as they have true zero measurements which represents no height, weight, speed, or time.

Ratio variables are the top level of measurement as you can order the observations, know the difference between any two values, and be able to assess ratios between values as they have a true zero measurement 

```{r quiz4, echo = FALSE}
quiz(caption = "Levels of Measurement Quiz",
  question("What is a nominal variable?",
    answer("Names of categories data can fall in", correct = TRUE),
    answer("Variables that can be ranked following a natural order of values"),
    answer("Variables measured on an interval scale"),
    answer("Variables that have a true zero measurement")
  ),
  question("What is an interval variable?",
    answer("Names of categories data can fall in"),
    answer("Variables that can be ranked following a natural order of values"),
    answer("Variables measured on an interval scale", correct = TRUE),
    answer("Variables that have a true zero measurement")
  ),
  question("Which of the following is an example of an ordinal variable?",
    answer("Gender"),
    answer("Eye color"),
    answer("Income level (low, middle, high)", correct = TRUE),
    answer("Height")
))


```

## Tasks
No tasks for this tutorial

### Reflect
Let's take some time to reflect before moving on. What is one thing that surprised you? What is one thing that confused you? Did you learn anything that might be useful in the type of work you do?

### Wrap-up
Good job on completing the first of two tutorials in the final module of this course! Please move onto the next tutorial within this module on Canvas where we will learn how generate descriptive statistics in R.


